Dish: Distribution-oblivious Shell Scripting

### 2019-11-25 pldi/draft.tex
Update section 3

Second, \sys needs to fill-in planing details left blank by the rewriting pass.
This is achieved with \emph{planner activators}, shell expressions that yield (and pass their arguments) back to the planner.
They are inserted by the planner and evaluated purely for their side-effects (\ie yielding to the scheduler) rather than their return value (which is the empty character).
They make clever use of shell features such as command expansion to achieve the latest possible binding, when critical runtime information is available.
% From the point of view of the shell, planner activators are expressed through a higher-order command---similar to \ttt{xargs} and \ttt{time}---that takes other commands as arguments.
Planner activators

@@

The next few sections (\S\ref{distributability}--\ref{impl}) discuss \sys's design and implementation details.
They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.

@@

To have any hope of distributing \unix pipelines, \sys first needs to understand the distribution characteristics of their intermediate commands.
This understanding is encoded in a command's \emph{distributability} class.
Informally, a distributability class represents the level of synchronization required by copies of a command executing in parallel.
It focuses on characteristics that are important for distribution, rather than all the observable behavior of a command.
\sys leans towards a few coarse classes rather than many detailed ones---among other reasons, to simplify their use by command developers.
% Make the connections that this is like a type, that doesn't capture everything, but is good-enough

This section starts by defining these classes, paired with a distributability study of the commands found in standard libraries such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
Building on this study, it presents a DSL that enables command classification by its developers~\sx{ext}.
In return, \sys uses this DSL to tag commands in the standard libraries and generate their wrappers, as presented in later sections.

@@

% nv TODO: this is a hierarchy
Broadly, shell commands can be split into five major classes (summarized in Tab.~\ref{tab:classes}) with respect to their distribution characteristics, depending on how they interact with state.
These classes are ordered in ascending difficulty of distribution:
  later classes require more effort.
In this order, each class can be thought as a subset of the next one---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).

@@

\heading{Stateless Commands}
The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
These are commands that can be expressed as a purely functional map---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert a {\sc NL} token.
Workloads that use only stateless commands are trivial to parallelize:
  they do not require any synchronization to maintain correctness, nor caution about where to split.
 % map :: (a -> b) -> [a] -> [b]

Many of these commands (about 1/3 of coreutils' \sta) are stateless \emph{within} an element (\eg \ttt{tr} transliterates characters, one at a time).
This property could allow further parallelization within a single line element, a feature that might seem of limited use because these commands are not particularly expensive computationally precisely due to their narrow focus.
However, it is useful for cases with very large stream elements (long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.

\heading{Pure Commands}
The second class contains pure commands \pur.
These commands respect functional purity, returning the same results for the same inputs, but maintain internal state across their entire pass.
The details of this state and its propagation during piecewise element processing affects their distributability characteristics.

@@

Often these commands do not operate in an online fashion, but need to block until the end of a stream---a typical example of this is \ttt{sort}, which cannot start before the last line of input has been consumed.
Non-streaming constraints affect task parallelism, but not data parallelism:
  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques, that is implementing it as a map and a reduce function.

\heading{File-system Commands}
The third class contains commands that access the file-system~\dfs.
This is the largest class, as the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
In fact, \unix uses the file-system as a proxy to several file-unrelated operations---\eg access control and device driving is grafted upon the file-system.

% TODO: mention that some commands fail to comply with the interface---for example, csplit could be split followed by a redirection to files.
Commands in \dfs are distributable for many reasons, as evidenced by a long history of distributed \unix file-system clones~\cite{catalogue}.
Some of these offer the semantics of the \unix file-system---hierarchies, protection bits---and a POSIX virtualization layer that enables support for the commands in \dfs.
The vast majority of these commands are control-plane operations that act on identifiers~\sx{bg}---for example, \ttt{touch} for creating a file or \ttt{chgrp} for changing group ownership.
% these operate by changing an inode
Data-plane operations on files go explicitly through stream-redirection operations~\sx{bg}.
Operations using different identifiers pose no problems;
  operations sharing a single identifier might have problems---but these risks exist even in the non-distributed case.
While such pathological cases are possible, they are rare enough---and have been rare enough since \unix's inception~\cite[\S3.6]{unix}).

@@

A command in \sid will require some form of concurrency control to distribute property.
However, because its side-effects stay within the distributed \unix environment, they can be manipulated using---including optimistic and multi-versioning schemes that can roll-back side-effects.

The vast majority of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never affected by the user.
For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information that is not writable by user-space scripts.
Even if \sid commands are protected with transactions
  these benefits are even greater for read-only or read-mostly accesses.

\heading{Irreversibly Side-effectful Commands}
Finally, some commands have side-effects outside \sys---for example, by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
Such commands are not distributable---they have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism.

\heading{Discussion}
There are several technical details about \sys's design decisions worth noting.

@@

Wanting to present the full picture for completeness purposes, the focus for the rest paper will be commands in \sta and \pur.
Other commands are treated as  non-distributable for now~\sx{discussion};
  by building infrastructure to handle other classes, \sys would be able to (i) extract more parallelism and (ii) handle larger programs.
Even with only these two classes, however, \sys supports highly useful pipelines that are more expressive than ones in popular frameworks such as Hadoop~\cite{mapreduce:08} and Spark~\cite{spark:12}~\sx{eval}.

>>>>

Second, \sys needs to fill-in planing details left blank by the rewriting pass.
This is achieved by \emph{planner activators}, shell expressions that yield (and pass their arguments) back to the planner.
They are inserted by the planner and evaluated purely for their side-effects (\ie yielding to the scheduler) rather than their return value (which is empty, not affecting pipelines).
They make clever use of shell features such as expansion to achieve the latest possible binding, when critical runtime information becomes available.
% From the point of view of the shell, planner activators are expressed through a higher-order command---similar to \ttt{xargs} and \ttt{time}---that takes other commands as arguments.

@@

The next few sections (\S\ref{distributability}--\ref{impl}) discuss \sys's design and implementation details.
% They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.

@@

To have any hope of distributing pipelines, \sys first needs to understand the distribution characteristics of their individual commands.
This understanding is encoded in a command's \emph{distributability} class.
Informally, a distributability class represents the level of synchronization required by copies of a command that execute in parallel.
It focuses on capturing characteristics that are important for distribution, rather than the entire the observable behavior of a command.
\sys leans towards having a few coarse classes rather than many detailed ones---among other reasons, to simplify their understanding and use by command developers.
% Make the connections that this is like a type, that doesn't capture everything, but is good-enough

This section starts by defining these classes, along with a distributability study of the commands found in standard libraries such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
Building on this study, it develops a DSL that enables command classification by its developers~\sx{ext}.
In return, \sys uses this DSL to tag commands in the standard libraries and generate their wrappers, as presented in later sections.

@@

% nv TODO: this is a hierarchy
Broadly, shell commands can be split into five major classes (summarized in Tab.~\ref{tab:classes}) with respect to their distribution characteristics, depending on how they interact with state.
These classes are ordered in ascending difficulty of distribution:
  later classes require more effort.
In this order, some classes can be thought as subsets of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).

@@

\heading{Stateless Commands}
The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert an {\sc NL} token.
Workloads that use only stateless commands are trivial to parallelize:
  they do not require any synchronization to maintain correctness, nor caution about where to be split.
 % map :: (a -> b) -> [a] -> [b]

Many of these commands (about 1/3 of coreutils' \sta) are stateless \emph{within} an element (\eg \ttt{tr} transliterates characters within a line, one at a time).
This property allows further parallelization, a feature that might seem of limited use as these commands are computationally inexpensive, precisely due to their narrow focus.
However, it turns to be  useful for cases with very large stream elements (long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.  % TODO pointer to 6.3

\heading{Pure Commands}
The second class contains pure commands \pur.
These commands respect functional purity---re\-turn\-ing the same results for the same inputs---but maintain internal state across their entire pass.
The details of this state and its propagation during piecewise element processing affect their distributability characteristics.

@@

Often these commands do not operate in an online fashion, but need to block until the end of a stream.
A typical example of this is \ttt{sort}, which cannot start emitting results before the last input element has been consumed.
Such constraints affect task parallelism, but not data parallelism:
  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---that is, by encoding it as a combination of a $map$ and a $reduce$ function.

\heading{File-system Commands}
The third class contains commands that access the file-system~\dfs.
This is the largest class, as the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
In fact, \unix uses the file-system as a proxy to several file-unrelated operations---\eg access control and device driving. % is grafted upon the file-system.

% TODO: mention that some commands fail to comply with the interface---for example, csplit could be split followed by a redirection to files.
Commands in \dfs are distributable for many reasons, as evidenced by a long history of distributed \unix file-system clones~\cite{catalogue}.
Some of these offer the semantics of the \unix file-system---hierarchies, protection bits---and a POSIX virtualization layer that enables support for all the commands in \dfs.
The vast majority of these commands are control-plane operations that act on identifiers~\sx{bg}---\eg \ttt{touch} for creating a file or \ttt{chgrp} for changing group ownership.
% these operate by changing an inode
Data-plane operations on files go explicitly through stream-redirection operations~\sx{bg}.
Operations using different identifiers pose no problems;
  operations sharing a single identifier might introduce problems---but these risks exist even in the non-distributed case.
Such pathological cases are indeed possible, but are rare enough to not pose a concern---and have been, since \unix's inception~\cite[\S3.6]{unix}.

@@

A command in \sid will require some form of concurrency control to distribute property.
However, because its side-effects stay within the distributed \unix environment, they can be manipulated using a transactional control mechanism---including optimistic and multi-versioning schemes that have the ability to roll-back side-effects.

The vast majority of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never written by user code.
For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information that is not writable by user-space scripts.
The benefits of a scalable transactional scheme are even greater when the workload features only or mostly read accesses.

\heading{Irreversibly Side-effectful Commands}
Finally, a few commands (about 0.4\%) have side-effects outside \unix---\eg by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
Such commands are not distributable---they often have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism that locks these resources.

\heading{Discussion}
There are a few technical details about \sys's design decisions worth noting.

@@

The focus for the rest paper will be commands in \sta and \pur.
\sys currently treats other commands as non-distributable~\sx{discussion}.
In future work, by building infrastructure to handle other classes, \sys will be able to (i) extract more parallelism and (ii) handle larger programs.
Even with only these two classes, however, \sys distributes highly useful pipelines that are strictly more expressive than programs written in popular frameworks such as Hadoop~\cite{mapreduce:08} and Spark~\cite{spark:12}~\sx{eval}.
